\textbf{Evaluation of machine learning methods for 16S rRNA gene data}

### 2019-03-07 submission/manuscript.Rmd
update text

Advances in sequencing technology and decreasing costs of generating 16S rRNA gene sequences have allowed rapid exploration of human associated microbiome and its health implications. Currently, the human microbiome field is growing at an unprecedented rate and as a result, there is an increasing demand for methods that identify associations between members of the microbiome and human health. However, this is an undertaking as human associated microbial communities are remarkably complex and uneven. It is unlikely that a single species can explain a disease. Instead, subsets of those communities, in relation to one another and to their host, account for the differences in the health outcomes. Therefore, researchers have started to explore the utility of machine learning (ML) models that use microbiota associated biomarkers to predict human health and to understand the microbial ecology of diseases such as liver cirrhosis, colorectal cancer, inflammatory bowel diseases (IBD), obesity, and type 2 diabetes [@zeller_potential_2014; @zackular_human_2014; @baxter_dna_2016; @baxter_microbiota-based_2016; @hale_shifts_2017; @pasolli_machine_2016; @sze_looking_2016; @walters_meta-analyses_2014; @vazquez-baeza_guiding_2018; @qin_alterations_2014; @geman_deep_2018]. ML methods are effective at recognizing and highlighting patterns in complex microbial datasets. However, currently the field's use of ML lacks clarity and consistency on which methods are used and how these methods are implemented. Moreover, there is a lack of deliberation on why a particular ML model is utilized. Recently, there is a trend towards using more complex ML models such as random forest, extreme gradient boosting and neural networks without a discussion on if and how much model interpretibility is necessary for the study [@galkin_human_2018; @reiman_using_2017; @fioravanti_phylogenetic_2017; @ @geman_deep_2018]. The lack of transparency on methodology and modeling decisions can have negative consequences on model reproducibility and reliability which we need avoid by (1) implementing consistent, accessible and transparent machine learning practices; (2) selecting ML models that reflect the goal of the study as it will inform the expectations of model accuracy, complexity, interpretibility and computational efficiency.

@@

We interpreted the feature weights of L1 and L2 SVM with linear kernel and regression coefficients of L2 logistic regression using the training data. To get a sense of the importance of each feature we randomly subsampled 80% of the data 100 times and trained 100 models; this resulted in 100 different sets of feature weights. We reported the mean weights of the 10 most important features [Figure 4]. FIT levels were

>>>>

Advances in sequencing technology and decreasing costs of generating 16S rRNA gene sequences have allowed rapid exploration of human associated microbiome and its health implications. Currently, the human microbiome field is growing at an unprecedented rate and as a result, there is an increasing demand for methods that identify associations between members of the microbiome and human health. However, this is an undertaking as human associated microbial communities are remarkably complex and uneven. It is unlikely that a single species can explain a disease. Instead, subsets of those communities, in relation to one another and to their host, account for the differences in the health outcomes. Therefore, researchers have started to explore the utility of machine learning (ML) models that use microbiota associated biomarkers to predict human health and to understand the microbial ecology of diseases such as liver cirrhosis, colorectal cancer, inflammatory bowel diseases (IBD), obesity, and type 2 diabetes [@zeller_potential_2014; @zackular_human_2014; @baxter_dna_2016; @baxter_microbiota-based_2016; @hale_shifts_2017; @pasolli_machine_2016; @sze_looking_2016; @walters_meta-analyses_2014; @vazquez-baeza_guiding_2018; @qin_alterations_2014; @geman_deep_2018]. ML methods are effective at recognizing and highlighting patterns in complex microbial datasets. However, currently the field's use of ML lacks clarity and consistency on which methods are used and how these methods are implemented. Moreover, there is a lack of deliberation on why a particular ML model is utilized. Recently, there is a trend towards using more complex ML models such as random forest, extreme gradient boosting and neural networks without a discussion on if and how much model interpretibility is necessary for the study [@galkin_human_2018; @reiman_using_2017; @fioravanti_phylogenetic_2017; @ @geman_deep_2018]. The lack of transparency on methodology and modeling decisions can have negative consequences on model reproducibility and reliability which we need to avoid by (1) implementing consistent, accessible and transparent machine learning practices; (2) selecting ML models that reflect the goal of the study as it will inform the expectations of model accuracy, complexity, interpretibility and computational efficiency.

@@

We interpreted the feature weights of L1 and L2 SVM with linear kernel and regression coefficients of L2 logistic regression using the training data. To get a sense of the importance of each feature we randomly subsampled 80% of the data 100 times and trained 100 models; this resulted in 100 different sets of feature weights. We reported the mean weights of the 10 most important features [Figure 4].

### 2019-03-07 submission/manuscript.tex
update text

To showcase transparent ML methodologies and to shed light on how much
ML model selection can affect modeling results, we performed an
empirical analysis comparing several different ML models using the same
dataset. We used a previously published colorectal cancer (CRC) study
(3) which had fecal 16S rRNA gene sequences and human hemoglobin
concentrations from 490 patients. We built ML models using fecal 16S
rRNA gene sequences and human hemoglobin concentrations to predict
patients with normal or screen relevant neoplasias (SRN) disease status.
The study had 261 normal and 229 SRN samples. We established modeling
pipelines for L2-regularized logistic regression, L1 and L2 support
vector machines (SVM) with linear and radial basis function kernels, a
decision tree, random forest and XGBoost which increase in complexity
and decrease in interpretability respectively. Our ML pipeline performed
100 data-splits and utilized held-out test data to evaluate
generalization and predicton performance of each ML model. The mean
AUROC varied from 0.68 (std ± 0.05) to 0.82 (std ± 0.04). Random Forest
and XGBoost had the highest mean AUROC for detecting SRN. Despite the
simplicity, the L1-regularized linear kernel SVM followed Random Forest
and XGBoost in performance. In terms of computational efficiency, L2 SVM
with linear kernel trained the fastest (0.127 hours, std ± 0.01), while
XGBoost took the longest (82.47 hours, std ± 17.045). We found that mean
cross-validation and testing AUROC could vary by as much as 0.061, which
highlights the importance of a seperate held-out test set for
evaluation. Aside from evaluating generalization and classification
performances for each of these models, this study established standards
for modeling pipelines of microbiome-associated machine learning models.

@@

We evaluated the prediction performance of seven binary classification
models when applied to held-out test data over 100 data-splits using
AUROC as classification performance metric. Random Forest and XGBoost
had the highest mean AUROC for detecting SRN, 0.816 (std ± 0.039) and
0.814 (std ± 0.04) respectively {[}Figure 2{]}. L1 linear SVM and
decision tree had significantly lower AUROC values, 0.75 (std ± 0.045)
and 0.741 (std ± 0.038) {[}Figure 2{]}. However, they had significanly
higher performances than L2 linear SVM, RBF SVM and L2 logistic
regression which had mean AUROC values of 0.68 (std ± 0.052), 0.684 (std
± 0.049) and 0.677 (std ± 0.053) respectively {[}Figure 2{]}. We also
evaluated the generalization performance of each classifier by comparing
their mean cross-validation AUROC and mean testing AUROC. We found a
statistically significant difference for classifiers L2 support vector
machines (SVM) with linear and radial basis function kernels and L2
logistic regression (p-values = 2.4e-16, 3.5e-13 and 9.4e-20
respectively). These differences were 0.052, 0.047 and 0.061,
respectively {[}Figure 2{]} .

@@

@@

Linear models trained faster than non-linear models. L2 logistic
Rregression and L1 and L2 SVM with linear kernel had training times of
7.8 min, (std ± 0.6), 9.6 min, (std ± 0.6) and 13.2 min, (std ± 1.2)
respectively. Whereas, SVM with radial basis function kernel, decision
tree, random forest and xgboost had training times of 2.21 hrs, (std ±
0.25), 4.2 hrs, (std ± 0.49), 82.47 hrs, (std ± 17.05) and 116.83 hrs,
(std ± 17.24), respectively.

>>>>

To showcase transparent ML methodologies and to shed light on how much
ML model selection can affect modeling results, we performed an
empirical analysis comparing several different ML models using the same
dataset. We used a previously published colorectal cancer (CRC) study
(3) which had fecal 16S rRNA gene sequences and human hemoglobin
concentrations from 490 patients. We built ML models using fecal 16S
rRNA gene sequences and human hemoglobin concentrations to predict
patients with normal or screen relevant neoplasias (SRN) disease status.
The study had 261 normal and 229 SRN samples. We established modeling
pipelines for L2-regularized logistic regression, L1 and L2 support
vector machines (SVM) with linear and radial basis function kernels, a
decision tree, random forest and XGBoost which increase in complexity
and decrease in interpretability respectively. Our ML pipeline performed
100 data-splits and utilized held-out test data to evaluate
generalization and predicton performance of each ML model. The mean
AUROC varied from 0.5 (std ± 0.01) to 0.82 (std ± 0.04). Random Forest
and XGBoost had the highest mean AUROC for detecting SRN. Despite the
simplicity, the L1-regularized linear kernel SVM followed Random Forest
and XGBoost in performance. In terms of computational efficiency, L2 SVM
with linear kernel trained the fastest (0.123 hours, std ± 0.014), while
XGBoost took the longest (106.568 hours, std ± 6.839). We found that
mean cross-validation and testing AUROC could vary by as much as 0.238,
which highlights the importance of a seperate held-out test set for
evaluation. Aside from evaluating generalization and classification
performances for each of these models, this study established standards
for modeling pipelines of microbiome-associated machine learning models.

@@

We evaluated the prediction performance of seven binary classification
models when applied to held-out test data over 100 data-splits using
AUROC as classification performance metric. Random Forest and XGBoost
had the highest mean AUROC for detecting SRN, 0.816 (std ± 0.039) and
0.814 (std ± 0.04) respectively {[}Figure 2{]}. L1 linear SVM and
decision tree had significantly lower AUROC values, 0.503 (std ± 0.027)
and 0.741 (std ± 0.038) {[}Figure 2{]}. However, they had significanly
higher performances than L2 linear SVM, RBF SVM and L2 logistic
regression which had mean AUROC values of 0.501 (std ± 0.013), 0.68 (std
± 0.05) and 0.672 (std ± 0.05) respectively {[}Figure 2{]}. We also
evaluated the generalization performance of each classifier by comparing
their mean cross-validation AUROC and mean testing AUROC. We found a
statistically significant difference for classifiers L2 support vector
machines (SVM) with linear and radial basis function kernels and L2
logistic regression (p-values = 2.4e-37, 1.3e-14 and 2.6e-21
respectively). These differences were 0.222, 0.05 and 0.064,
respectively {[}Figure 2{]} .

@@

We interpreted the feature weights of L1 and L2 SVM with linear kernel
and regression coefficients of L2 logistic regression using the training
data. To get a sense of the importance of each feature we randomly
subsampled 80\% of the data 100 times and trained 100 models; this
resulted in 100 different sets of feature weights. We reported the mean
weights of the 10 most important features {[}Figure 4{]}.

@@

Linear models trained faster than non-linear models. L2 logistic
Rregression and L1 and L2 SVM with linear kernel had training times of
7.2 min, (std ± 0.6), 10.2 min, (std ± 0.6) and 13.2 min, (std ± 1.8)
respectively. Whereas, SVM with radial basis function kernel, decision
tree, random forest and xgboost had training times of 2.01 hrs, (std ±
0.28), 4.2 hrs, (std ± 0.49), 106.57 hrs, (std ± 6.84) and 116.83 hrs,
(std ± 17.24), respectively.
